The dissertation is based on the following papers. They are presented in the order of publication.
\subsubsection{Paper 1}
EM Pedersen, E Agerbo, O Plana-Ripoll, J Grove, JW Dreier, KL Musliner, M Bækvad-Hansen, G Athanasiadis, A Schork, D Demontis, J Bybjerg-Grauholm, DM Hougaard, T Werge, M Nordentoft, O Mors, S Dalsgaard, J Christensen,  AD Børglum, PB Mortensen, JJ McGrath, F Privé, BJ Vilhjálmsson. Accounting for age-of-onset and family history improves power in genome-wide association studies. American Journal of Human Genetics, 109: 417-432.

\subsubsection{Paper 2}
EM Pedersen, E Agerbo, O Plana-Ripoll, J Steinbach, MD Krebs, DM Hougaard, T Werge, M Nordentoft, A Børglum,  KL Musliner, A Ganna, AJ Schork, PB Mortensen,  JJ McGrath,  F Privé, BJ Vilhjálmsson. ADuLT: An efficient and robust time-to-event GWAS. medRxiv, doi: https://doi.org/10.1101/2022.08.11.22278618 [Under review]

\subsubsection{Paper 3}
\textbf{\textit{Study 3:}} fGRS and fGRS multi trait [TBD - Under construction]
\mbox{}\\
Besides these I have contributed to several other manuscripts that are not included in this dissertation. This includes the three published studies and one pre-print.

\begin{enumerate}
	\item MJ Witteveen, EM Pedersen, J Meijsen, MR Andersen, F Privé, D Speed, and BJ Vilhjálmsson. Publicly Available Privacy-preserving Benchmarks for Polygenic Prediction. bioRxiv, \url{https://doi.org/10.1101/2022.10.10.510645}
	\item T Wimberley, I Brikell, EM Pedersen, E Agerbo, BJ Vilhjálmsson, C Albiñana, F Privé, A Thapar, K Langley, L Riglin, M Simonsen, HS Nielsen, AD Børglum, M Nordentoft, PB Mortensen, S Dalsgaard. Early life injuries and the development of attention-deficit hyperactivity disorder.  Journal of Clinical Psychiatry, doi:\url{https://doi.org/10.4088/JCP.21m14033}.
	\item I Brikell, T Wimberley, C Albiñana, EM Pedersen, BJ Vilhjálmsson, E Agerbo, D Demontis, AD Børglum, A Schork, S LaBianca, T Werge, M Nordentoft, O Mors, D Hougaard, A Thapar, PB Mortensen, S Dalsgaard. Genetic, Clinical, and Sociodemographic Factors Associated With Stimulant Treatment Outcomes in ADHD.  American Journal of Psychiatry, doi:\url{https://doi.org/10.1176/appi.ajp.2020.20121686}.
	\item X Liu, T Munk-Olsen, C Albiñana, BJ Vilhjálmsson, E Pedersen, V Schlünssen, M Bækvad-Hansen, J Bybjerg-Grauholm, M Nordentoft, A Børglum, T Werge, D Hougaard, PB Mortensen, E Agerbo. Genetic liability to major depression and risk of childhood asthma. Brain Behavior and Immunity, doi: \url{https://doi.org/10.1016/j.bbi.2020.07.030}.
	
\end{enumerate}
