


This section will summarise the results of the papers the dissertation is based on. All papers will utilise some version of the age-dependent liability threshold model. Each paper has its own distinct use case of the model, which will be highlighted in the coming sections.

\section{Paper 1 - LT-FH++}
\input{results/ltfhpp_text.tex}
\newpage

\section{Paper 2 - ADuLT}
\input{results/adult_text.tex}
\newpage

\section{Paper 3 - fGRS}
TBA
%\input{results/}




perhaps structure the results section by paper in this way:

\begin{enumerate}
	\item LT-FH++: highlight Mortality and ADHD. It shows the importance of including family history and age-of-onset to increase power, but also shows that it is not the end all be all.
	\begin{enumerate}
		\item Mortality GWAS - large power increase
		\item ADHD GWAS - almost no power increase
	\end{enumerate}
	\item ADuLT: when only looking at age-of-onset, cox PH models are probably not the best model to use. ADuLT is \textit{never} the worst model, but also not always the best. The most robust method is ADuLT, since power is always best or close to the best.
	\begin{enumerate}
		\item simulation results with downsampling
		\item ADHD GWAS results
	\end{enumerate}
	\item fGRS: the predictive performance of the liabilities compared to binary variables.
	\begin{enumerate}
		\item single trait performance
		\item multi trait performance 
		\item cross ancestry performance
	\end{enumerate}	
	\item Do I need a section combining the results into a larger picture somehow here ? or is it meant to go in the discussion?
\end{enumerate}
